MSSM-Penn-Pitt - ACC and DLPFC - FEATURECOUNTS - GRCh38 - CQN - Iterative Model

### 2020-05-07 code/CMC_ExpVal_DataProcessing.Rmd
Update validationcode

                    block = COVARIATES1$Block, 
                    correlation = correlation$cor)
# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS$E,
            weights = VOOM.WEIGHTS$weights,
            design = VOOM.WEIGHTS$design[,colnames(VOOM.NET.FIT)],
            block = COVARIATES1$Block, 
            correlation = correlation$cor)

# Get (null) design matrix
MODEL0 = DESIGN$design[,-grep(primaryVariable[2],colnames(DESIGN$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E)
# Compute actual variance of all principal components
expr = RESIDUAL.SVA.GENE_EXPRESSION
expr[is.na(expr)] = 0
tmp = svd(expr)
actual.var = tmp$d^2/sum(tmp$d^2)
# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(expr, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])
# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p
# Estimate surrogate variables
SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E.no.na,
                        MODEL1, MODEL0, n.sv = NUM.SV, 
                        B = 30, tol = 1e-20)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)
# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(cnts, design=cbind(MODEL1, SURR.VAR), plot=F)
# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(cnts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES1$Block, 
                        correlation = correlation$cor)
# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS$E,
                     weights = VOOM.SVA.WEIGHTS$weights,
                     design = VOOM.SVA.WEIGHTS$design,
                     block = COVARIATES1$Block, 
                     correlation = correlation$cor)
# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E)
# Add variable of interest back to the residuals
varsToAddIn = grep("Tissue.Diagnosis", colnames(MODEL1), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```

```{r EXP_Val, include=FALSE, eval=TRUE, cache=FALSE}
################################################################################
#CMC ACC DATA
library(stringr)

@@

  setEPS()
  postscript(paste0( Tissue, "_Cor_gt5Perc.eps"), onefile = TRUE)
    hist( Correlation[Correlation$FDR_W_Only < 0.05, ]$Correlation, 
        breaks = 100, 
        xlab = 'Correlation',
        col = 'grey45',
        las = 1,
        main="Correlation of Genes with FDR Corrected P-Value < 0.05")
  dev.off()
  Filt_Hist <- hist( Correlation[Correlation$FDR_W_Only < 0.05, ]$Correlation, 
        breaks = 100, 
        xlab = 'Correlation',
        col = 'grey45',
        las = 1,
        main="Correlation of Genes with FDR Corrected P-Value < 0.05")
  return(list( TAB=Tab, Sig = Sig, Tot=Total_Hist, Filt=Filt_Hist ))
}

ACC_Plots <- Associated( 'ACC' )
PFC_Plots <- Associated( 'PFC' )

@@

>>>>

                    block = COVARIATES1$Individual_ID, 
                    correlation = correlation$cor)
# Fit linear model using new weights and new design
VOOM.WEIGHTS$E = CQN.GENE_EXPRESSION$E
FIT = lmFit(VOOM.WEIGHTS$E,
            weights = VOOM.WEIGHTS$weights,
            design = VOOM.WEIGHTS$design[,colnames(VOOM.NET.FIT)],
            block = COVARIATES1$Individual_ID, 
            correlation = correlation$cor)

# Get (null) design matrix
MODEL0 = DESIGN$design[,-grep(primaryVariable[2],colnames(DESIGN$design))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN$design
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]
# Get residuals from differential expression model
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(FIT, CQN.GENE_EXPRESSION$E)
# Compute actual variance of all principal components
expr = RESIDUAL.SVA.GENE_EXPRESSION
expr[is.na(expr)] = 0
tmp = svd(expr)
actual.var = tmp$d^2/sum(tmp$d^2)
# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind) %dopar% {
  tmp.residual = t(apply(expr, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])
# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p
# Estimate surrogate variables
SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E.no.na,
                        MODEL1, MODEL0, n.sv = NUM.SV, 
                        B = 30, tol = 1e-20)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)
# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(cnts, design=cbind(MODEL1, SURR.VAR), plot=F)
# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(cnts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F,
                        block = COVARIATES1$Individual_ID, 
                        correlation = correlation$cor)
# Fit linear model using new weights and new design
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS$E,
                     weights = VOOM.SVA.WEIGHTS$weights,
                     design = VOOM.SVA.WEIGHTS$design,
                     block = COVARIATES1$Individual_ID, 
                     correlation = correlation$cor)
# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E)
# Add variable of interest back to the residuals
varsToAddIn = grep("Tissue.Diagnosis", colnames(MODEL1), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```

```{r EXP_Val, include=FALSE, eval=TRUE, cache=FALSE}
################################################################################

#CMC ACC DATA
library(stringr)

@@

  setEPS()
  postscript(paste0( Tissue, "_Cor_gt5Perc.eps"), onefile = TRUE)
    hist( Correlation[Correlation$FDR_W_Only < 0.05, ]$Correlation, 
        breaks = 100, 
        xlab = 'Correlation',
        col = 'grey45',
        las = 1,
        main="Correlation of Genes with FDR Corrected P-Value < 0.05")
  dev.off()
  Filt_Hist <- hist( Correlation[Correlation$FDR_W_Only < 0.05, ]$Correlation, 
        breaks = 100, 
        xlab = 'Correlation',
        col = 'grey45',
        las = 1,
        main="Correlation of Genes with FDR Corrected P-Value < 0.05")
  return(list( TAB=Tab, Correlation=Correlation, Sig = Sig, Tot=Total_Hist, Filt=Filt_Hist ))
}

ACC_Plots <- Associated( 'ACC' )
PFC_Plots <- Associated( 'PFC' )
#save.image('Completed_Validation_Analysis.RData')
```

@@

#ENRICH_OBJ <-  syn_temp$store( synapseclient$File( path='Completed_Validation_Analysis.RData', name = 'RData object of analysis for ACC Validation', parentId=CODE$properties$id ), activityName = activityName, executed = c( Syns_Used, thisFile) , activityDescription = activityDescription)
#  all.annotations$tissueTypeAbrv = 'Analysis'
#  syn_temp$setAnnotations(ENRICH_OBJ, annotations = all.annotations)
